---
title: "MinneMUDAC"
author: "Makenna Halvensleben"
date: '2023-02-27'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(corrplot)
library(lubridate)
library(data.table)
library(tidymodels)
library(ranger)
```

# Reading in data

```{r}
sched23 <- read_csv("2023_MLBSchedule.csv")
scheds <- read_csv("OriginalSchedules.csv",
                   col_types = cols(DateofMakeup = col_character()),
                   locale = locale(encoding = "UTF-16LE"))
logs <- read_csv("GameLogs.csv", 
                 col_types = cols(Completition_Information = col_character(), 
                                  Forfeit_Information = col_character(), 
                                  Protest_Information = col_character(),
                                  LFUmp_ID = col_character(),
                                  RFUmp_ID = col_character(),
                                  Additional_Information = col_character()), 
                 locale = locale(encoding = "UTF-16LE"))


logs$Date <- ymd(logs$Date)
logs$Year <- year(logs$Date)

logs <- logs %>% mutate_at(vars(contains("_Position")),
                               funs(case_when(.=="1"~"Pitcher",
                                              .=="2"~"Catcher",
                                              .=="3"~"First",
                                              .=="4"~"Second",
                                              .=="5"~"Third",
                                              .=="6"~"Short",
                                              .=="7"~"Left",
                                              .=="8"~"Center",
                                              .=="9"~"Right")))
```

# Dealing with NA

Factors with mostly NA values are removed from the data set

```{r}
na_count <- sapply(logs, function(y) sum(length(which(is.na(y)))))

na_count <- data.frame(na_count)

clean_logs <- logs %>% select(-c(Forfeit_Information, Protest_Information, LFUmp_ID, RFUmp_ID, Completition_Information, Additional_Information, SavingPitcher_ID))
```

# Dealing with 0 Attendance

Entries with 0 or negative attendance are removed from the data set

```{r}
nrow(clean_logs[which(clean_logs$Attendance <= 0),])
clean_logs <- clean_logs %>% filter(Attendance > 0)
```

# Creating data frame with just MIN data

```{r}
min_logs <- clean_logs %>% filter(HomeTeam == "MIN")
```

# Looking at Attendance distribution

## Just MIN

```{r}
min_logs %>% ggplot() + geom_histogram(aes(x = Attendance)) + xlab("Attendance for MN Twins") + theme_bw()
min_logs %>% ggplot() + geom_boxplot(aes(x = Attendance))
summary(min_logs$Attendance)
```

## All Teams

```{r}
clean_logs %>% ggplot() + geom_histogram(aes(x = clean_logs$Attendance)) + xlab("Attendance for All Teams") + theme_bw()
clean_logs %>% ggplot() + geom_boxplot(aes(x = Attendance))
summary(clean_logs$Attendance)
```

# Identifying important factors

## Just MIN

### Corrplot with top 5 negative correlation and top 5 positive correlation

Correlation plots are only valid for numeric factors, so only those are considered.

Correlation plot is organized so that the 5 factors with the greatest positive correlation are listed first followed by the 5 factors with the greatest negative correlation.

```{r}
min_numeric <- min_logs[,sapply(min_logs, is.numeric)]
min_cor <- cor(min_numeric)
min_order <- min_cor[order(min_cor[,"Attendance"]),]
min_names <- row.names(min_order)
min_cor_top10 <- min_numeric %>% select(c(min_names[c(1:5,60:65)])) %>% rev() %>% cor()
corrplot(min_cor_top10, tl.cex = 0.5, type="full", method = "color")
```

#### Bar plots of important factors

```{r}
min_logs %>% ggplot() + geom_bar(aes(x = VisitingTeamGameNumber, y = Attendance), stat="identity")
min_logs %>% ggplot() + geom_bar(aes(x = HomeTeamGameNumber, y = Attendance), stat="identity")
min_logs %>% ggplot() + geom_bar(aes(x = HomeTeamDefense_Assists, y = Attendance), stat="identity")
min_logs %>% ggplot() + geom_bar(aes(x = VisitingTeamDefense_Assists, y = Attendance), stat="identity")
min_logs %>% ggplot() + geom_bar(aes(x = Year, y = Attendance), stat="identity")

min_logs %>% ggplot() + geom_bar(aes(x = VisitingTeamOffense_Strickouts, y = Attendance), stat="identity")
min_logs %>% ggplot() + geom_bar(aes(x = VisitingTeamOffense_AwardedFirstonCatcherInterference, y = Attendance), stat="identity")
min_logs %>% ggplot() + geom_bar(aes(x = VisitingTeamPitchers_PitchersUsed, y = Attendance), stat="identity")
min_logs %>% ggplot() + geom_bar(aes(x = VisitingTeamOffense_Homeruns, y = Attendance), stat="identity")
min_logs %>% ggplot() + geom_bar(aes(x = HomeTeamOffense_AwardedFirstonCatcherInterference, y = Attendance), stat="identity")
```

### Stepwise Model

```{r echo=FALSE}
min.null <- lm(Attendance ~ .,data=min_numeric)
min.step <- stats::step(min.null)
```
```{r}
summary(min.step)
```

## All MLB Teams

### Corrplot with top 5 negative correlation and top 5 positive correlation

Correlation plots are only valid for numeric factors, so only those are considered.

Correlation plot is organized so that the 5 factors with the greatest positive correlation are listed first followed by the 5 factors with the greatest negative correlation.

```{r}
all_numeric <- clean_logs[,sapply(clean_logs, is.numeric)]
all_cor <- cor(all_numeric)
all_order <- all_cor[order(all_cor[,"Attendance"]),]
all_names <- row.names(all_order)
all_cor_top10 <- all_numeric %>% select(c(all_names[c(1:5,60:65)])) %>% rev() %>% cor()
corrplot(all_cor_top10, tl.cex = 0.7, type = "full", method = "color")
```

#### Bar plots of important factors

```{r}
clean_logs %>% ggplot() + geom_bar(aes(x = VisitingTeamGameNumber, y = Attendance), stat="identity")
clean_logs %>% ggplot() + geom_bar(aes(x = HomeTeamGameNumber, y = Attendance), stat="identity")
clean_logs %>% ggplot() + geom_bar(aes(x = LengthofGame, y = Attendance), stat="identity")
clean_logs %>% ggplot() + geom_bar(aes(x = HomeTeamOffense_SacrificeHits, y = Attendance), stat="identity")
clean_logs %>% ggplot() + geom_bar(aes(x = HomeTeamOffense_IntentionalWalks, y = Attendance), stat="identity")

clean_logs %>% ggplot() + geom_bar(aes(x = Year, y = Attendance), stat="identity")
clean_logs %>% ggplot() + geom_bar(aes(x = NumberofGames, y = Attendance), stat="identity")
clean_logs %>% ggplot() + geom_bar(aes(x = HomeTeamOffense_Strickouts, y = Attendance), stat="identity")
clean_logs %>% ggplot() + geom_bar(aes(x = HomeTeamPitchers_WildPitches, y = Attendance), stat="identity")
# spelled visiting wrong
clean_logs %>% ggplot() + geom_bar(aes(x = VistingTeamScore, y = Attendance), stat="identity")
```

## Stepwise Model

```{r echo = FALSE}
all.null <- lm(Attendance ~ .,data=all_numeric)
all.step <- stats::step(all.null)
```
```{r}
summary(all.step)
```

# Predicting attendance for all teams using random forest

## Using all of the available factors

### Splitting data into training and testing

```{r}
set.seed(1234)

attendance_split <- clean_logs %>%
  initial_split(strata = HomeTeam)

train <- training(attendance_split)
train <- na.omit(train)
train <- train %>% mutate_if(is.factor, as.character)
test <- testing(attendance_split)
test <- na.omit(test)
test <- test %>% mutate_if(is.factor, as.character)

test$isTest <- rep(1,nrow(test))
train$isTest <- rep(0,nrow(train))
fullSet <- rbind(test,train)
fullSet <- fullSet %>% mutate_if(is.character, as.factor)
test.new <- fullSet[fullSet$isTest==1,]
train.new <- fullSet[fullSet$isTest==0,]

rf_spec_ranger <- rand_forest(mode = "regression") %>%
  set_engine("ranger")
```

### Fitting random forest model

```{r}
rf_fit <- rf_spec_ranger %>%
  fit(Attendance ~ .,
    data = train.new
  )

rf_fit
```

### Predicting with training and testing data

```{r}
pred <- rf_fit %>% predict(new_data = test.new)
pred

results_train <- rf_fit %>%
    predict(new_data = train.new) %>%
    mutate(
      truth = train.new$Attendance,
      model = "rf"
    )

results_test <- rf_fit %>%
    predict(new_data = test.new) %>%
    mutate(
      truth = test.new$Attendance,
      model = "rf"
    )

results_train %>%
  group_by(model) %>%
  rmse(truth = truth, estimate = .pred)

results_test %>%
  group_by(model) %>%
  rmse(truth = truth, estimate = .pred)

results_test %>%
  mutate(train = "testing") %>%
  bind_rows(results_train %>%
    mutate(train = "training")) %>%
  ggplot(aes(truth, .pred, color = model)) +
  geom_abline(lty = 2, color = "gray80", size = 2) +
  geom_point(alpha = 0.5) +
  facet_wrap(~train) +
  labs(
    x = "Truth",
    y = "Predicted attendance",
    color = "Type of model"
  ) + theme_bw(base_size = 15)

results_test %>%
  mutate(train = "testing") %>%
  ggplot(aes(truth, .pred, color = model)) +
  geom_abline(lty = 2, color = "gray80", size = 2) +
  geom_point(alpha = 0.5) +
  facet_wrap(~train) +
  labs(
    x = "Truth",
    y = "Predicted attendance",
    color = "Type of model"
  ) + theme_bw(base_size = 15)
```

## Using just the factors that can be deduced from the 2023 schedule data

### Splitting data into training and testing

```{r}
clean_logs_known <- clean_logs %>% select(Date, DayofWeek, VisitingTeam, VisitingTeamLeague, VisitingTeamGameNumber, HomeTeam, HomeTeamLeague, HomeTeamGameNumber, DayNight, BallParkID, Year, Attendance)

attendance_split_known <- clean_logs_known %>%
  initial_split(strata = HomeTeam)

train_known <- training(attendance_split_known)
train_known <- na.omit(train_known)
train_known <- train_known %>% mutate_if(is.factor, as.character)
test_known <- testing(attendance_split_known)
test_known <- na.omit(test_known)
test_known <- test_known %>% mutate_if(is.factor, as.character)

test_known$isTest <- rep(1,nrow(test_known))
train_known$isTest <- rep(0,nrow(train_known))
fullSet_known <- rbind(test_known,train_known)
fullSet_known <- fullSet_known %>% mutate_if(is.character, as.factor)
test.new_known <- fullSet_known[fullSet_known$isTest==1,]
train.new_known <- fullSet_known[fullSet_known$isTest==0,]
```

### Fitting random forest model

```{r}
rf_fit_known <- rf_spec_ranger %>%
  fit(Attendance ~ .,
    data = train.new_known
  )

rf_fit_known
```

### Predicting with training and testing data

```{r}
pred_known <- rf_fit_known %>% predict(new_data = test.new_known)
pred_known


results_train_known <- rf_fit_known %>%
    predict(new_data = train.new_known) %>%
    mutate(
      truth = train.new_known$Attendance,
      model = "rf"
    )

results_test_known <- rf_fit_known %>%
    predict(new_data = test.new_known) %>%
    mutate(
      truth = test.new_known$Attendance,
      model = "rf"
    )

results_train_known %>%
  group_by(model) %>%
  rmse(truth = truth, estimate = .pred)

results_test_known %>%
  group_by(model) %>%
  rmse(truth = truth, estimate = .pred)

results_test_known %>%
  mutate(train = "testing") %>%
  bind_rows(results_train_known %>%
    mutate(train = "training")) %>%
  ggplot(aes(truth, .pred, color = model)) +
  geom_abline(lty = 2, color = "gray80", size = 2) +
  geom_point(alpha = 0.5) +
  facet_wrap(~train) +
  labs(
    x = "Truth",
    y = "Predicted attendance",
    color = "Type of model"
  ) + theme_bw(base_size = 15)

results_test_known %>%
  mutate(train = "testing") %>%
  ggplot(aes(truth, .pred, color = model)) +
  geom_abline(lty = 2, color = "gray80", size = 2) +
  geom_point(alpha = 0.5) +
  facet_wrap(~train) +
  labs(
    x = "Truth",
    y = "Predicted attendance",
    color = "Type of model"
  ) + theme_bw(base_size = 15)
```

### Adding factors that can be deduced to the 2023 schedule data set

Rare stadiums are removed from the data set. These include Estadio Alfredo Harp Helu, London Stadium, and Muncy Bank Ballpark.

```{r}
sched23$Date <- ymd(sched23$game_date)
sched23$Year <- year(sched23$Date)
sched23$DayofWeek <- wday(sched23$Date)
sched23$DayofWeek <- case_when(
  sched23$DayofWeek==1~"Mon",
  sched23$DayofWeek==2~"Tue",
  sched23$DayofWeek==3~"Wed",
  sched23$DayofWeek==4~"Thu",
  sched23$DayofWeek==5~"Fri",
  sched23$DayofWeek==6~"Sat",
  sched23$DayofWeek==7~"Sun")
sched23$HomeTeam <- sched23$home_team
sched23$VisitingTeam <- sched23$away_team
sched23$DayNight <- ifelse(sched23$game_time < hms("17:00:00"), "D", "N")
sched23$VisitingTeamLeague <- ""
sched23$HomeTeamLeague <- ""
sched23$VisitingTeamGameNumber <- 0
sched23$HomeTeamGameNumber <- 0

sched23 <- sched23 %>% filter(stadium_name != "Estadio Alfredo Harp Helu", stadium_name != "London Stadium", stadium_name != "Muncy Bank Ballpark")

sched23$BallParkID <- case_when(
  sched23$HomeTeam=="ANA"~"ANA01",
  sched23$HomeTeam=="ARI"~"PHO01",
  sched23$HomeTeam=="ATL"~"ATL03",
  sched23$HomeTeam=="BAL"~"BAL12",
  sched23$HomeTeam=="BOS"~"BOS07",
  sched23$HomeTeam=="CHA"~"CHI12",
  sched23$HomeTeam=="CHN"~"CHI11",
  sched23$HomeTeam=="CIN"~"CIN09",
  sched23$HomeTeam=="CLE"~"CLE08",
  sched23$HomeTeam=="COL"~"DEN02",
  sched23$HomeTeam=="DET"~"DET05",
  sched23$HomeTeam=="HOU"~"HOU03",
  sched23$HomeTeam=="KCA"~"KAN06",
  sched23$HomeTeam=="LAN"~"LOS03",
  sched23$HomeTeam=="MIA"~"MIA02",
  sched23$HomeTeam=="MIL"~"MIL06",
  sched23$HomeTeam=="MIN"~"MIN04",
  sched23$HomeTeam=="NYA"~"NYC21",
  sched23$HomeTeam=="NYN"~"NYC20",
  sched23$HomeTeam=="OAK"~"OAK01",
  sched23$HomeTeam=="PHI"~"PHI13",
  sched23$HomeTeam=="PIT"~"PIT08",
  sched23$HomeTeam=="SDN"~"SAN02",
  sched23$HomeTeam=="SEA"~"SEA03",
  sched23$HomeTeam=="SFN"~"SFO03",
  sched23$HomeTeam=="SLN"~"STL10",
  sched23$HomeTeam=="TBA"~"STP01",
  sched23$HomeTeam=="TEX"~"ARL03",
  sched23$HomeTeam=="TOR"~"TOR02",
  sched23$HomeTeam=="WAS"~"WAS11")

for (x in 1:2425) {
  sched23[x,]$VisitingTeamLeague <- logs[which(logs$VisitingTeam == sched23[x,]$VisitingTeam),][1,]$VisitingTeamLeague
  sched23[x,]$HomeTeamLeague <- logs[which(logs$HomeTeam == sched23[x,]$HomeTeam),][1,]$HomeTeamLeague
}

sched23 <- sched23 %>% mutate_if(is.character, as.factor)

gameNumbers <- as.data.frame(matrix(nrow=30, ncol=2))
gameNumbers$V1 <- unique(sched23$HomeTeam)
gameNumbers$V2 <- 1

for (x in 1:2425) {
  visiting <- gameNumbers[which(gameNumbers$V1 == sched23[x,]$VisitingTeam),]$V2
  home <- gameNumbers[which(gameNumbers$V1 == sched23[x,]$HomeTeam),]$V2
  sched23[x,]$VisitingTeamGameNumber <- visiting
  sched23[x,]$HomeTeamGameNumber <- home
  gameNumbers[which(gameNumbers$V1 == sched23[x,]$VisitingTeam),]$V2 <- visiting + 1
  gameNumbers[which(gameNumbers$V1 == sched23[x,]$HomeTeam),]$V2 <- home + 1
}

sched23$isTest <- 0
```

### Predicting with 2023 schedule data

```{r}
pred_sched23 <- rf_fit_known %>% predict(new_data = sched23)
pred_sched23

sched23$pred <- pred_sched23$.pred
```

## Using all data to train random forest when predicting for 2023 schedule

### Fitting random forest model

```{r}
rf_fit_all_data <- rf_spec_ranger %>%
  fit(Attendance ~ .,
    data = clean_logs_known
  )

rf_fit_all_data
```

### Predicting with 2023 schedule

```{r}
pred_sched23_all <- rf_fit_all_data %>% predict(new_data = sched23)
pred_sched23_all

sched23$pred_all <- pred_sched23_all$.pred
```

## Comparing predictions

```{r}
pred_sched23 %>% ggplot() + geom_histogram(aes(x = .pred))
summary(pred_sched23$.pred)
pred_sched23_all %>% ggplot() + geom_histogram(aes(x = .pred))
summary(pred_sched23_all$.pred)
```

# Predicting attendance for all teams using linear regression

```{r}
clean_logs_linear <- clean_logs %>% select(Date, DayofWeek, VisitingTeam, VisitingTeamLeague, VisitingTeamGameNumber, HomeTeam, HomeTeamLeague, HomeTeamGameNumber, DayNight, Year, Attendance)

attendance_split_linear <- clean_logs_linear %>%
  initial_split(strata = HomeTeam)

train_linear <- training(attendance_split_linear)
train_linear <- na.omit(train_linear)
train_linear <- train_linear %>% mutate_if(is.factor, as.character)
test_linear <- testing(attendance_split_linear)
test_linear <- na.omit(test_linear)
test_linear <- test_linear %>% mutate_if(is.factor, as.character)

test_linear$isTest <- rep(1,nrow(test_linear))
train_linear$isTest <- rep(0,nrow(train_linear))
fullSet_linear <- rbind(test_linear,train_linear)
fullSet_linear <- fullSet_linear %>% mutate_if(is.character, as.factor)
test.new_linear <- fullSet_linear[fullSet_linear$isTest==1,]
train.new_linear <- fullSet_linear[fullSet_linear$isTest==0,]

lm_spec <- linear_reg(mode = "regression") %>%
  set_engine("glm")

lm_fit <- lm_spec %>%
  fit(Attendance ~ .,
    data = train.new_linear
  )

lm_fit

predlm <- lm_fit %>% predict(new_data = test.new_linear)
predlm

results_train_linear <- lm_fit %>%
    predict(new_data = train.new_linear) %>%
    mutate(
      truth = train.new_linear$Attendance,
      model = "lm"
    )

results_test_linear <- lm_fit %>%
    predict(new_data = test.new_linear) %>%
    mutate(
      truth = test.new_linear$Attendance,
      model = "lm"
    )

results_train_linear %>%
  group_by(model) %>%
  rmse(truth = truth, estimate = .pred)

results_test_linear %>%
  group_by(model) %>%
  rmse(truth = truth, estimate = .pred)

results_test_linear %>%
  mutate(train = "testing") %>%
  bind_rows(results_train_linear %>%
    mutate(train = "training")) %>%
  ggplot(aes(truth, .pred, color = model)) +
  geom_abline(lty = 2, color = "gray80", size = 2) +
  geom_point(alpha = 0.5) +
  facet_wrap(~train) +
  labs(
    x = "Truth",
    y = "Predicted attendance",
    color = "Type of model"
  ) + theme_bw(base_size = 15)

results_test_linear %>%
  mutate(train = "testing") %>%
  ggplot(aes(truth, .pred, color = model)) +
  geom_abline(lty = 2, color = "gray80", size = 2) +
  geom_point(alpha = 0.5) +
  facet_wrap(~train) +
  labs(
    x = "Truth",
    y = "Predicted attendance",
    color = "Type of model"
  ) + theme_bw(base_size = 15)

summary(lm_fit$fit)

```

## Comparing random forest with linear regression

```{r}
results_train_compare <- lm_fit %>%
  predict(new_data = train.new_linear) %>%
  mutate(
    truth = train.new_linear$Attendance,
    model = "lm"
  ) %>%
  bind_rows(rf_fit_known %>%
    predict(new_data = train.new_known) %>%
    mutate(
      truth = train.new_known$Attendance,
      model = "rf"
    ))

results_test_compare <- lm_fit %>%
  predict(new_data = test.new_linear) %>%
  mutate(
    truth = test.new_linear$Attendance,
    model = "lm"
  ) %>%
  bind_rows(rf_fit_known %>%
    predict(new_data = test.new_known) %>%
    mutate(
      truth = test.new_known$Attendance,
      model = "rf"
    ))

results_test_compare %>%
  mutate(train = "testing") %>%
  bind_rows(results_train_compare %>%
    mutate(train = "training")) %>%
  ggplot(aes(truth, .pred, color = model)) +
  geom_abline(lty = 2, color = "gray80", size = 2) +
  geom_point(alpha = 0.5) +
  facet_wrap(~train) +
  labs(
    x = "Truth",
    y = "Predicted attendance",
    color = "Type of model"
  ) + theme_bw(base_size = 15)

results_test_compare %>%
  mutate(train = "testing") %>%
  ggplot(aes(truth, .pred, color = model)) +
  geom_abline(lty = 2, color = "gray80", size = 2) +
  geom_point(alpha = 0.5) +
  facet_wrap(~train) +
  labs(
    x = "Truth",
    y = "Predicted attendance",
    color = "Type of model"
  ) + theme_bw(base_size = 15)
```

## Metrics

```{r}
rf_fit_known
rf_fit_all_data
summary(lm_fit$fit)
glance(lm_fit)
```

## Writing to csv

```{r}
write.csv(sched23, "sched23pred.csv")
```

# Various Notes

summary

residuals

confusion matrix

post hoc analysis of random forest model


1 lm     regression
2 glm    regression
3 glmnet regression
4 stan   regression
5 spark  regression
6 keras  regression
7 brulee regression


coefficient

confint

relevel

glm

Look at categorical variables?????

Random Forrest

https://www.listendata.com/2014/11/random-forest-with-r.html#:~:text=Yes%2C%20it%20can%20be%20used,numeric%20or%20continuous%20dependent%20variable. 

Gradient Boosting Tree

https://www.projectpro.io/recipes/apply-gradient-boosting-r-for-regression

look at the one lady's code

association

https://stackoverflow.com/questions/52554336/plot-the-equivalent-of-correlation-matrix-for-factors-categorical-data-and-mi
